---
title: "Dynamic Data"
author: "Ian Lyttle"
date: "`r Sys.Date()`"
output: github_document
---

The purpose of this document is to download the datasets used from dynamic sources, in particular the lastest data from the Iowa Coronavirus website.

For now, this has to be run interactively because I can't get promises to play well with knitting RMarkdown files.

```{r}
library("fs")
library("crrri")
```


```{r}
dir_create("data")

dir_target <- path("data", "download")
dir_create(dir_target)
```

The data from the NYT is delayed by a day. It may be useful to scrape the Iowa Coronavirus site for the latest information.

```{r}
url_access <- "https://coronavirus.iowa.gov/pages/access"
html_file <- path(dir_target, "access.html")
css_selector <- "td"

snapshot <- file_info(html_file <- path(dir_target, "access.html"))
```

We are going to use [rvest](https://rvest.tidyverse.org/) to page, but there's a problem. The page is not a static HTML site; the data we want is injected into the site using JavaScript. This means that **rvest** would be looking for the css tags in a document that does not have the data table.

We need another tool for the toolbox: [crrri](https://rlesur.github.io/crrri) - a package written by Romain Lesur and Christophe Dervieux - which will let us build the page locally, using Chrome. Then we will scrape it, using our local copy of the page.

The **crrri** package is not on CRAN, and it requires that you have Chrome installed on your computer, so that it can call it in the background. To run Chrome, the **crrri** package will need to know where to *find* Chrome. You can use `pagedown::find_chrome()` to find the location on your computer (on mine, it is `"/Applications/Google Chrome.app/Contents/MacOS/Google Chrome"`). The **crrri** package uses the environment variable `HEADLESS_CHROME` to look for a default value.

```{r}
chrome <- Chrome$new()
client <- chrome$connect()
```

```{r}
dump_DOM <- function(client, url, html_file) {
  Network <- client$Network
  Page <- client$Page
  Runtime <- client$Runtime
  Network$enable() %...>%
  { Page$enable() } %...>%
  { Network$setCacheDisabled(cacheDisabled = TRUE) } %...>% 
  { Page$navigate(url = url) } %...>%
  { Page$loadEventFired() } %...>% {
    Sys.sleep(10) # hacky - wait for a certain event, instead
  } %...>% { 
    Runtime$evaluate(
      expression = 'document.documentElement.outerHTML'
    ) 
  } %...>% {
    writeLines(c(.$result$value, "\n"), con = html_file) 
  } %>%
  finally(
    ~ client$disconnect()
  ) %...!% {
    cat("Error:", .$message, "\n")
  }
}
```

```{r}
client %...>% 
  dump_DOM(url_access, html_file)
```

```{r}
client
```



