---
title: "COVID-19 Iowa"
author: "Ian Lyttle"
date: "`r Sys.Date()`"
output: 
  github_document:
    fig_width: 9 
    fig_height: 4.5 
params:
  threshold_cases: 50 # for individual counties
  threshold_index_cases: 20 # for indexing FT-style charts
  threshold_new_cases: 5
  threshold_index_new_cases: 5
  nyt_citation: "Data from NYT, via state and local health agencies"
  iowa_citation: "Data from Iowa Community Indicators Program, via US Census Bureau"
---

```{r}
library("fs")
library("glue")
library("tidyverse")
library("USAboundaries") # also install_github("ropensci/USAboundariesData")
library("sf")
library("colorspace")
```

```{r}
dir_source <- path("data", "wrangle")
dir_target <- path("data", "iowa")

dir_create(dir_target)
```

## Prepare data

Let's read in the county data:

```{r}
iowa_counties <- 
  read_csv(path(dir_source, "iowa-counties.csv")) %>%
  print()
```

Let's have a quick look at all the counties.

```{r county-raw}
ggplot(iowa_counties, aes(date, cases)) +
  geom_line(aes(color = county)) +
  scale_y_log10()
```

This plot is not all that informative, except to tell us that we have the data we expect.

Let's get just the data for the most-current data.

```{r}
iowa_counties_current <- 
  iowa_counties %>%
  filter(date == max(date)) %>%
  arrange(desc(cases)) %>%
  print()
```

It could be useful to see the current distribution of number of cases per county:

```{r county-histogram}
date_max <- unique(iowa_counties_current$date)

ggplot(iowa_counties_current) +
  geom_histogram(aes(x = cases), binwidth = 2) +
  theme_bw() + 
  labs(
    x = "number of reported cases in county",
    y = "number of counties",
    title = glue("Iowa: distribution of counties by number of COVID-19 cases - {date_max}"),
    caption = params$nyt_citation  
  ) 
```

On the right, we see Johnson, Polk, and Linn counties, with the greatest number of reported cases. On the left, we see a lot of counties with only a few cases. 

Our eyes will let us pay attention to only a finite number of things at a time, so I propose to set a threshold of `r params$threshold_cases` cases. We identify those counties where the number of cases is greater than or equal to the threshold.

```{r}
counties_large <- 
  iowa_counties_current %>%
  filter(cases >= params$threshold_cases) %>%
  arrange(desc(cases)) %>%
  pull(county)

counties_large
```

In addition to compiliing the data for the counties with large numbers of cases, we also create aggregeted datasets that show:

- the total for the entire state.
- the total for all the counties that are not considered individually.

```{r}
iowa_total <- 
  iowa_counties %>%
  group_by(date) %>%
  summarize(
    county = factor(NA_character_, levels = counties_large),
    cases = sum(cases),
    deaths = sum(deaths),
    new_cases = sum(new_cases),
    new_deaths = sum(new_deaths),
    new_cases_week_avg = sum(new_cases_week_avg),
    new_deaths_week_avg = sum(new_deaths_week_avg),
    aggregation = "total"
  ) %>%
  arrange(desc(date)) %>%
  print()
```

```{r}
iowa_remainder <- 
  iowa_counties %>%
  filter(!(county %in% counties_large)) %>%
  group_by(date) %>%
  summarize(
    county = factor(NA_character_, levels = counties_large),
    cases = sum(cases),
    deaths = sum(deaths),
    new_cases = sum(new_cases),
    new_deaths = sum(new_deaths),
    new_cases_week_avg = sum(new_cases_week_avg),
    new_deaths_week_avg = sum(new_deaths_week_avg),
    aggregation = "remainder"
  ) %>%
  arrange(desc(date)) %>%
  print()
```

```{r}
iowa_counties_large <- 
  iowa_counties %>%
  filter(county %in% counties_large) %>%
  mutate(
    county = factor(county, levels = counties_large)
  )
```

Let's combine our datasets:

```{r}
data_combined <- bind_rows(iowa_counties_large, iowa_total, iowa_remainder)
```


If we wanted do one of those [FT-style charts](https://www.ft.com/coronavirus-latest), we need to choose an index day for each county and aggregation. We choose an arbitrary threshold of `r params$threshold_index_cases` reported cases.

```{r}
data_combined_index <- 
  data_combined %>%
  group_by(aggregation, county) %>%
  arrange(date) %>%
  mutate(
    exceeds_threshold_cases = cases >= params$threshold_index_cases,
    index_cases = cumsum(exceeds_threshold_cases) - 1,
    exceeds_threshold_new_cases = new_cases >= params$threshold_new_cases,
    index_new_cases = cumsum(cumany(exceeds_threshold_new_cases)) - 1
  ) 
```

We combine all these datasets into a plot:

```{r}
plot_cases <- 
  ggplot(data_combined, aes(date, cases)) +
  geom_line(aes(color = county, linetype = aggregation)) +
  geom_point(
    data = . %>% filter(date == max(date)),
    aes(color = county),
    show.legend = FALSE
  ) +
  scale_y_log10(sec.axis = dup_axis(name = NULL)) + 
  scale_linetype(
    name = "grouping", 
    breaks = c("total", "remainder", "none"),
    labels = c("state total", "all other counties", "county")
  ) +
  scale_color_discrete(breaks = counties_large) +
  theme_bw() + 
  labs(
    x = NULL,
    y = "number of reported cases",
    title = "Iowa: COVID-19 cases by reporting date",
    subtitle = "Includes counties with most reported cases, also aggregations",
    caption = params$nyt_citation  
  )
```

```{r}
plot_new_cases_week_avg <- 
  ggplot(data_combined, aes(date, new_cases_week_avg)) +
  geom_line(aes(color = county, linetype = aggregation)) +
  geom_point(
    data = . %>% filter(date == max(date)),
    aes(color = county),
    show.legend = FALSE
  ) +
  scale_y_log10(sec.axis = dup_axis(name = NULL)) + 
  scale_linetype(
    name = "grouping", 
    breaks = c("total", "remainder", "none"),
    labels = c("state total", "all other counties", "county")
  ) +
  scale_color_discrete(breaks = counties_large) +
  theme_bw() + 
  labs(
    x = NULL,
    y = "number of new reported cases",
    title = "Iowa: COVID-19 daily new cases (7-day rolling average) by reporting date",
    subtitle = "Includes counties with most reported cases, also aggregations",
    caption = params$nyt_citation  
  )
```

```{r}
# function that takes a set of limits, returns the breaks
# - we want breaks every 7 days.
f_break <- function(limits) {

  low <- floor(limits[1] / 7) * 7
  hi <- ceiling(limits[2] / 7) * 7
  
  seq(low, hi, by = 7)
}

# function that returns data for "doubling" lines
#' @param data          `data.frame`
#' @param var_index     `character`, name of variable in `data` containing day-index
#' @param var_number    `character`, name of variable in `data` contiaining number-data
#' @param days_doubling `numeric`, vector with number-of-days to include
#'
#' @return `data.frame` with variables `days_doubling`, `index`, `number`, `label`
#'
f_doubling <- function(data, var_index, var_number, threshold_number,
                       days_doubling = c(2, 4, 7, 14)) {

  index_range <- range(data[[var_index]])
  number_range <- range(data[[var_number]])
  
  max_number <- number_range[2]
  max_index <- index_range[2]

  # for each `days_doubling`, determine the index when days are maximum 
  by_number <- 
    crossing(days_doubling, number = max_number) %>%
    mutate(
      index = days_doubling * log(number/threshold_number, 2)
    )
    
  by_index <- 
    crossing(days_doubling, index = index_range) %>%
    mutate(
      number = threshold_number * 2^(index/days_doubling)
    )

  combined <- 
    dplyr::bind_rows(by_index, by_number) %>%
    dplyr::select(days_doubling, index, number) %>%
    dplyr::mutate(
      label = ifelse(index > 0, glue::glue("{days_doubling}-day doubling"), "")
    ) %>%
    dplyr::filter(index <= max_index, number <= max_number) %>%
    dplyr::arrange(days_doubling, index) 
  
  combined
}

plot_cases_ft <- 
  ggplot(
    data_combined_index %>% filter(index_cases >= 0), 
    aes(index_cases, cases)
  ) +
  geom_line(
    data = 
      . %>% 
      f_doubling("index_cases", "cases", params$threshold_index_cases), 
    aes(x = index, y = number, group = factor(days_doubling)),
    color = "grey80",
    linetype = "twodash"
  ) +
  geom_text(
    data = 
      . %>% 
      f_doubling("index_cases", "cases", params$threshold_index_cases) %>%
      filter(str_length(label) > 0), 
    aes(x = index, y = number, group = factor(days_doubling), label = label),
    hjust = "right", 
    nudge_y = 0.05,
    color = "grey80"
  ) +
  geom_line(aes(color = county, linetype = aggregation)) +
  geom_point(
    data = . %>% filter(date == max(date)),
    aes(color = county),
    show.legend = FALSE
  ) +
  geom_text(
    data = . %>% filter(date == max(date), aggregation == "none"),
    aes(x = index_cases, label = county, color = county),
    hjust = "left",
    nudge_x = .5,
    size = 3,
    show.legend = FALSE
  ) +
  geom_text(
    data = . %>% 
      ungroup() %>%
      filter(date == max(date), aggregation != "none") %>%
      mutate(
        label = fct_recode(
          aggregation, 
          `state total` = "total", 
          `all other counties` = "remainder"
        )
      ),
    aes(x = index_cases, label = label),
    color = "grey10",
    hjust = "left",
    nudge_x = 0.5,
    size = 3,
    show.legend = FALSE
  ) +
  scale_x_continuous(breaks = f_break, expand = expand_scale(add = c(0, 3))) +
  scale_y_log10(
    sec.axis = dup_axis(name = NULL)
  ) + 
  scale_linetype(
    name = "grouping", 
    breaks = c("total", "remainder", "none"),
    labels = c("state total", "all other counties", "county")
  ) +  
  scale_color_discrete(breaks = counties_large) +
  guides(color = FALSE, linetype = FALSE) +
  theme_bw() +
  labs(
    x = glue("days since {params$threshold_index_cases} cases"),
    y = "number of reported cases",
    title = glue("Iowa: COVID-19 cumulative-case trajectories as of {max(data_combined_index$date)}"),
    subtitle = "Includes counties with most reported cases, also aggregations",
    caption = params$nyt_citation
  )
```

```{r}
plot_new_cases_week_avg_ft <- 
  ggplot(
    data_combined_index %>% filter(index_new_cases >= 0),
    aes(index_new_cases, new_cases_week_avg)
  ) +
  geom_line(aes(color = county, linetype = aggregation)) +
  geom_point(
    data = . %>% filter(date == max(date)),
    aes(color = county),
    show.legend = FALSE
  ) +
  geom_text(
    data = . %>% filter(date == max(date), aggregation == "none"),
    aes(x = index_new_cases, label = county, color = county),
    hjust = "left",
    nudge_x = .5,
    size = 3,
    show.legend = FALSE
  ) +
  geom_text(
    data = . %>% 
      ungroup() %>%
      filter(date == max(date), aggregation != "none") %>%
      mutate(
        label = fct_recode(
          aggregation, 
          NULL = "county",
          `state total` = "total", 
          `all other counties` = "remainder"
        )
      ),
    aes(x = index_new_cases, label = label),
    color = "grey10",
    hjust = "left",
    nudge_x = 0.5,
    size = 3,
    show.legend = FALSE
  ) +
  scale_x_continuous(breaks = f_break, expand = expand_scale(add = c(0, 3.5))) +
  scale_y_log10(
    sec.axis = dup_axis(name = NULL)
  ) + 
  scale_linetype(
    name = "grouping", 
    breaks = c("total", "remainder", "none"),
    labels = c("state total", "all other counties", "county")
  ) +
  scale_color_discrete(breaks = counties_large) +
  guides(color = FALSE, linetype = FALSE) +
  theme_bw() +
  labs(
    x = glue("days since {params$threshold_index_new_cases} new cases"),
    y = "number of new reported cases",
    title = glue("Iowa: COVID-19 cumulative-case trajectories as of {max(data_combined_index$date)}"),
    subtitle = "Includes counties with most reported cases, also aggregations",
    caption = params$nyt_citation
  )
```

### County groups

Let's look at the poulation of counties. I'd like to sort Iowa's counties into four groups, ordered by county-population, such that each *group* of counties contains approximately the same population. 

```{r}
iowa_county_population <- 
  read_csv(path(dir_source, "iowa-county-population.csv")) %>%
  print()
```

You can see that the "first" group contains two counties: Polk and Linn; the "second" group contains seven counties, from Scott to Dallas. This second group of counties is home to the Regents' Universities, Iowa's part of the Quad cities, Dubuque, Sioux City, and suburban Des Moines.

The "third" and "fourth" groups contain 24 and 66 counties, respectively. Here's a map of Iowa's counties, showing these groups:

```{r}
iowa_map <- us_counties(resolution = "low", state = "Iowa")
```

I am interested to compare how the trajectory of recorded cases behaves, or might behave differently, among these groups of counties.

```{r}
group_colors <- 
  colorspace::sequential_hcl(palette = "Blues 3", n = 6)[1:4] # drop lightest

iowa_map_with_group <- 
  iowa_map %>%
  left_join(
    iowa_county_population %>% select(county, population_group),
    by = c(name = "county")  
  )
```

```{r}
plot_iowa_map_group <-
  ggplot() +
  geom_sf(
    data = iowa_map_with_group, 
    aes(fill = population_group),
    color = "white"
  ) + 
  scale_fill_manual(values = group_colors) + 
  labs(
    title = "Iowa counties by population-group",
    subtitle = "Each population-group has about 1/4 of state's population (2019 estimate)",
    caption = params$iowa_citation,
    fill = "group"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    axis.text = element_blank()
  )
```

Let's look at proportions of counties, population, and (most-recent) reported cases and deaths.

```{r}
iowa_counts <-
  iowa_county_population %>% select(county, population, population_group) %>%
  left_join(
    iowa_counties_current,
    by = "county"
  ) %>%
  mutate(
    cases = ifelse(is.na(cases), 0, cases),
    deaths = ifelse(is.na(deaths), 0, deaths)
  ) %>%
  group_by(population_group) %>%
  summarize(
    counties = n(),
    population = sum(population),
    cases = sum(cases),
    deaths = sum(deaths)
  ) %>%
  print()
```

```{r}
vars_count <- c("counties", "population", "cases", "deaths")

iowa_counts_tall <-
  iowa_counts %>%
  pivot_longer(
    all_of(vars_count), 
    names_to = "category", 
    values_to = "count"
  ) %>%
  mutate(
    category = factor(category, rev(vars_count))
  ) %>%
  print()
```

```{r}
plot_iowa_proportions <-
  ggplot(
    iowa_counts_tall, 
    aes(x = category, y = count)
  ) + 
  geom_col(aes(fill = population_group), position = position_fill(reverse = TRUE)) + 
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = group_colors) + 
  coord_flip() +
  labs(
    x = "",
    y = "proportion",
    fill = "group",
    title = glue("Iowa: COVID-19 proportions - as of {date_max}"),
    caption = params$nyt_citation
  ) +
  theme_bw()
```

```{r fig.width=9, fig.height=4.5}
iowa_counties_group <-
  iowa_counties %>%
  left_join(
    iowa_county_population %>% select(county, population_group),
    by = "county"
  )

iowa_counties_group_agg <-
  iowa_counties_group %>%
  group_by(date, population_group) %>%
  summarize(
    cases = sum(cases),
    deaths = sum(deaths),
    new_cases = sum(new_cases),
    new_deaths = sum(new_deaths),
    new_cases_week_avg = sum(new_cases_week_avg),
    new_deaths_week_avg = sum(new_deaths_week_avg)
  ) %>%
  ungroup()
```


```{r}
plot_cases_group <- 
  ggplot(iowa_counties_group_agg, aes(date, cases)) +
  geom_line(
    aes(color = population_group)
  ) +  
  geom_point(
    data = . %>% filter(date == max(date)),
    aes(color = population_group),
    show.legend = FALSE
  ) +
  scale_color_manual(values = group_colors) +
  scale_y_log10(sec.axis = dup_axis(name = NULL)) + 
  theme_bw() + 
  labs(
    x = NULL,
    y = "number of reported cases",
    color = "group",
    title = "Iowa: COVID-19 cases by reporting date",
    caption = params$nyt_citation  
  )
```

```{r}
plot_new_cases_group <- 
  ggplot(iowa_counties_group_agg, aes(date, new_cases_week_avg)) +
  geom_line(
    aes(color = population_group)
  ) +  
  geom_point(
    data = . %>% filter(date == max(date)),
    aes(color = population_group),
    show.legend = FALSE
  ) +
  scale_color_manual(values = group_colors) +
  scale_y_log10(sec.axis = dup_axis(name = NULL)) + 
  theme_bw() + 
  labs(
    x = NULL,
    y = "number of new reported cases",
    color = "group",
    title = "Iowa: COVID-19 daily new cases (7-day rolling average) by reporting date",
    caption = params$nyt_citation  
  )
```

```{r}
iowa_counties_group_agg_ft <- 
  iowa_counties_group_agg %>%
  group_by(population_group) %>%
  arrange(date) %>%
  mutate(
    exceeds_threshold_cases = cases > params$threshold_index_cases,
    index_cases = cumsum(exceeds_threshold_cases) - 1,
    exceeds_threshold_new_cases = new_cases >= params$threshold_index_new_cases,
    index_new_cases = cumsum(cumany(exceeds_threshold_new_cases)) - 1    
  ) 

plot_cases_group_ft <-
  ggplot(
    iowa_counties_group_agg_ft %>% filter(exceeds_threshold_cases), 
    aes(index_cases, cases)) +
  geom_line(
    data = 
      . %>% 
      f_doubling("index_cases", "cases", params$threshold_index_cases), 
    aes(x = index, y = number, group = factor(days_doubling)),
    color = "grey80",
    linetype = "twodash"
  ) +
  geom_text(
    data = 
      . %>% 
      f_doubling("index_cases", "cases", params$threshold_index_cases) %>%
      filter(str_length(label) > 0), 
    aes(x = index, y = number, label = label),
    hjust = "right", 
    nudge_y = 0.05,
    color = "grey80"
  ) +
  geom_line(aes(color = population_group)) +
  geom_point(
    data = . %>% filter(date == max(date)),
    aes(color = population_group),
    show.legend = FALSE
  ) +
  scale_x_continuous(breaks = f_break) +
  scale_y_log10(
    sec.axis = dup_axis(name = NULL)
  ) + 
  scale_color_manual(values = group_colors) +
  theme_bw() +
  labs(
    x = glue("days since {params$threshold_index_cases} cases"),
    y = "number of reported cases",
    color = "group",
    title = "Iowa: COVID-19 cumulative-case trajectories",
    caption = params$nyt_citation
  )
```

```{r}
plot_new_cases_group_ft <-
  ggplot(
    iowa_counties_group_agg_ft %>% filter(index_new_cases >= 0), 
    aes(index_new_cases, new_cases_week_avg)) +
  geom_line(aes(color = population_group)) +
  geom_point(
    data = . %>% filter(date == max(date)),
    aes(color = population_group),
    show.legend = FALSE
  ) +
  scale_x_continuous(breaks = f_break) +
  scale_y_log10(
    sec.axis = dup_axis(name = NULL)
  ) + 
  scale_color_manual(values = group_colors) +
  theme_bw() +
  labs(
    x = glue("days since {params$threshold_index_new_cases} new cases"),
    y = "number of new reported cases",
    color = "group",
    title = "Iowa: COVID-19 new-case trajectories (7-day rolling average)",
    caption = params$nyt_citation
  )
```

## Discussion

```{r cases-county}
plot_cases
```

For another view, here's an [FT-style chart](https://www.ft.com/coronavirus-latest):

```{r cases-county-ft}
plot_cases_ft
```

```{r new-cases-county}
plot_new_cases_week_avg
```

```{r new-cases-county-ft}
plot_new_cases_week_avg_ft
```

```{r iowa-map}
plot_iowa_map_group
```

```{r iowa-proportions}
plot_iowa_proportions
```

```{r cases-group}
plot_cases_group
```

```{r new-cases-group}
plot_new_cases_group
```

```{r cases-group-ft}
plot_cases_group_ft
```
```{r new-cases-group-ft}
plot_new_cases_group_ft
```
